---
title: "MDA 9159 - Statistical Modelling 1 - Fall 2024"
author: "Group - Team Bits: Miles Xi, Pengwei Xu, Lingyu Zhao"
output:
  pdf_document: default
  word_document: default
  html_document: default
urlcolor: blue
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r, include = FALSE}
library(knitr)
#opts_chunk$set(tidy.opts=list(width.cutoff=70),tidy=TRUE)
```

\newpage

# Loading Date and Data Exploratory Analysis

```{r}
library(dplyr)

# Read in dataset
data = read.csv("Team_Bits_Data.csv")

# Remove price=0 entries
data = data[data$PRICE != 0, ]

# Remove rows with NA in all columns except 'YR_RMDL'
# will impute the missing values, if any, for this variable
data = data %>% filter(!if_any(-YR_RMDL, is.na))

# Remove redundant columns 
# as "_D" columns are merely descriptions of their corresponding columns
data = data %>% select(-SSL, -OBJECTID, -GIS_LAST_MOD_DTTM, 
                       -QUALIFIED, -SALE_NUM, -BLDG_NUM,
                       -STYLE_D, -STRUCT_D, -GRADE_D,
                       -CNDTN_D, -EXTWALL_D, -ROOF_D,
                       -INTWALL_D, -USECODE, -HEAT_D,
                       -NUM_UNITS, -STRUCT)

head(data)
```

### Variable Explanation

We are dealing with housing data in this report, below is an explanation of variables:

  1. PRICE: response
  2. BATHRM: # bathrooms
  3. HF_BATHRM: # half bathrooms
  4. HEAT: heating
  5. AC: air conditioning
  6. ROOMS: # rooms
  7. BEDRM: # bedrooms
  8. AYB: The earliest time the main portion of the building was built
  9. YR_RMDL: Year structure was remodelled
  10. EYB: The year an improvement was built 
  11. STORIES: # stories in primary dwelling
  12. SALEDATE: Date of sale
  13. GBA: Gross building area in square feet
  14. STYLE: House style
  15. GRADE: House grade
  16. CNDTN: House condition
  17. EXTWALL: Exterior wall tyle
  18. ROOF: Roof type
  19. INTWALL: Interior wall type
  20. KITCHENS: # kitchens
  21. FIREPLACES: # fireplaces
  22. LANDAREA: Land area of property in square feet


### NA Data

Now let us explore the percentage of missing data for each predictor:

```{r}
missing_data = round(sapply(data, function(x) mean(is.na(x) * 100)), 3)

missing_data
```

From the R output above, observe that "YR_RMDL: Year structure was remodeled" has around 36% missing data. A possible explanation for this is that not all buildings were remodeled.


### Preprocessing
* Converted some predictors to numerical values:

  * AC: "Y" and "N" corresponds to "1" and "0".
  
  * SALEDATE: Transform calendar format values in SALEDATE to numerical values using as.Date().
  
* Created dummy variables for categorical predictors:

  * These categorical variables include: "HEAT", "STYLE", "GRADE", "CNDTN", "EXTWALL", "ROOF", "INTWALL", and "AC".

* Introduced a few new variables:

  * `SALE_YEAR`: The year that the house was sold, it is derived from `SALEDATE`.
  
  * `SALE_AYB_DIFF`: The difference between the year sold and the year built.
  
  * `SALE_EYB_DIFF`: The difference between the year sold and the year an improvement was applied.

  * `SALE_RMDL_DIFF`: The difference between the year sold and the year structure was remodeled.'
  

```{r}
library(lubridate)
library(fastDummies)

# Transform Yes/No for having AC to numerical values
data$AC = ifelse(data$AC == 'Y', 1, 0)

# Create dummy variables for categorical predictors
data = dummy_cols(
  data, 
  select_columns = c("HEAT", "STYLE", "GRADE", "CNDTN", 
                     "EXTWALL", "ROOF", "INTWALL", "AC"), 
  remove_selected_columns = TRUE,
  remove_first_dummy = TRUE
)

# Add SALEYEAR
data$SALE_YEAR = year(ymd_hms(data$SALEDATE))

# Add SALEYEAR and AYB diff
data$SALE_AYB_DIFF = data$SALE_YEAR - data$AYB

# Add SALEYEAR and EYB diff
data$SALE_EYB_DIFF = data$SALE_YEAR - data$EYB

# Add SALEYEAR and YR_RMDL diff
data$SALE_RMDL_DIFF = data$SALE_YEAR - data$YR_RMDL

# Convert SALEDATE column to numeric values
data$SALEDATE = as.numeric(as.Date(data$SALEDATE))
```

```{r}
# Define box-cox and inverse box-cox transformation
powerfun = function(y, lambda) {
  if (lambda == 0) {
    return(log(y))
  } else {
    return((y^lambda - 1) / lambda)
  }
}

inv_powerfun = function(y_transformed, lambda) {
  if (lambda == 0) {
    return(exp(y_transformed))
  } else {
    return((lambda * y_transformed + 1)^(1/lambda))
  }
}
```

### Data for Training and Validating

```{r}
set.seed(9159)

# Randomly sample 600 data entries for our project
clean_data = data[sample(nrow(data), 600),]

data_train = clean_data[1:500, ]  # First 500 rows for training
data_valid = clean_data[501:600, ]  # Last 100 rows for validation
```

```{r}
# Replace NA with column median
data_train = data.frame(lapply(data_train, function(column) {
  column_median = median(column, na.rm = TRUE)
  column[is.na(column)] = column_median
  column
}))

data_test = data.frame(lapply(data_test, function(column) {
  column_median = median(column, na.rm = TRUE)
  column[is.na(column)] = column_median
  column
}))
```