---
title: "LASSO"
output:
  pdf_document: default
  word_document: default
  html_document: default
urlcolor: blue
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r, include = FALSE}
library(knitr)
#opts_chunk$set(tidy.opts=list(width.cutoff=70),tidy=TRUE)
```

\newpage

# Loading Date and Data Exploratory Analysis

```{r}
library(dplyr)

# Read in dataset
data = read.csv("Team_Bits_Data.csv")

# Remove price=0 entries
data = data[data$PRICE != 0, ]

# Remove rows with NA in all columns except 'YR_RMDL'
data = data %>% filter(!if_any(-YR_RMDL, is.na))

# Remove not useful columns
data = data %>% select(-SSL, -OBJECTID, -GIS_LAST_MOD_DTTM, 
                       -QUALIFIED, -SALE_NUM, -BLDG_NUM,
                       -STYLE_D, -STRUCT_D, -GRADE_D,
                       -CNDTN_D, -EXTWALL_D, -ROOF_D,
                       -INTWALL_D, -USECODE, -HEAT_D,
                       -NUM_UNITS, -STRUCT)

head(data)
```

## Variable Explanation

We are dealing with housing data in this report, let me go over through the meanings behind each predictor:

  1. PRICE: response
  2. BATHRM: # bathrooms
  3. HF_BATHRM: # half bathrooms
  4. HEAT: heating
  5. AC: air conditioning
  6. ROOMS: # rooms
  7. BEDRM: # bedrooms
  8. AYB: The earliest time the main portion of the building was built
  9. YR_RMDL: Year structure was remodelled
  10. EYB: The year an improvement was built 
  11. STORIES: # stories in primary dwelling
  12. SALEDATE: Date of sale
  13. GBA: Gross building area in square feet
  14. STYLE: House style
  15. GRADE: House grade
  16. CNDTN: House condition
  17. EXTWALL: Exterior wall tyle
  18. ROOF: Roof type
  19. INTWALL: Interior wall type
  20. KITCHENS: # kitchens
  21. FIREPLACES: # fireplaces
  22. LANDAREA: Land area of property in square feet


## NA Data

Now let us explore the percentage of missing data for each predictor:

```{r}
missing_data = round(sapply(data, function(x) mean(is.na(x) * 100)), 3)

missing_data
```

From the R output above, observe that "YR_RMDL: Year structure was remodeled" has around 36% missing data. A possible explanation for this could be: not all buildings were remodeled.


## Preprocessing
* Converted some predictors to numerical values:

  * AC: "Y" and "N" corresponds to "1" and "0".
  
  * SALEDATE: Transform calendar format values in SALEDATE to numerical values using as.Date().
  
* Created dummy variables for categorical predictors:

  * These categorical variables include: "HEAT", "STYLE", "GRADE", "CNDTN", "EXTWALL", "ROOF" and "INTWALL".

* Introduced a few new variables:

  * `SALE_YEAR`: The year that the house was sold, it is derived from `SALEDATE`.
  
  * `SALE_AYB_DIFF`: The difference between the year sold and the year built.
  
  * `SALE_EYB_DIFF`: The difference between the year sold and the year an improvement was applied.

  * `SALE_RMDL_DIFF`: The difference between the year sold and the year structure was remodeled.'
  

```{r}
library(lubridate)
library(fastDummies)

# Transform Yes/No for having AC to numerical values
data$AC = ifelse(data$AC == 'Y', 1, 0)

# Create dummy variables for categorical predictors
data = dummy_cols(
  data, 
  select_columns = c("HEAT", "STYLE", "GRADE", "CNDTN", 
                     "EXTWALL", "ROOF", "INTWALL", "AC"), 
  remove_selected_columns = TRUE,
  remove_first_dummy = TRUE
)

# Add SALEYEAR
data$SALE_YEAR = year(ymd_hms(data$SALEDATE))

# Add SALEYEAR and AYB diff
data$SALE_AYB_DIFF = data$SALE_YEAR - data$AYB

# Add SALEYEAR and EYB diff
data$SALE_EYB_DIFF = data$SALE_YEAR - data$EYB

# Add SALEYEAR and YR_RMDL diff
data$SALE_RMDL_DIFF = data$SALE_YEAR - data$YR_RMDL

# Convert SALEDATE column to numeric values
data$SALEDATE = as.numeric(as.Date(data$SALEDATE))
```


```{r}
# Replace NA with column median
data = data.frame(lapply(data, function(column) {
  column_median = median(column, na.rm = TRUE)
  column[is.na(column)] = column_median
  column
}))
```


```{r}
# Define box-cox and inverse box-cox transformation
powerfun = function(y, lambda) {
  if (lambda == 0) {
    return(log(y))
  } else {
    return((y^lambda - 1) / lambda)
  }
}

inv_powerfun = function(y_transformed, lambda) {
  if (lambda == 0) {
    return(exp(y_transformed))
  } else {
    return((lambda * y_transformed + 1)^(1/lambda))
  }
}
```

## Data for Training and Validating

```{r}
set.seed(9159)

# Randomly sample 600 data entries for our project
clean_data = data[sample(nrow(data), 600),]

data_train = clean_data[1:500, ]  # First 500 rows for training
data_valid = clean_data[501:600, ]  # Last 100 rows for validation
```

## Data Visualization

```{r}
library(lmtest)
```

```{r}
# Plot histogram for untransformed response variable `PRICE`
hist(clean_data$PRICE, col=adjustcolor('firebrick',alpha=0.6),
     xlab='PRICE($)', ylab='Frequency',
     main='Histogram of Untransformed PRICE')
```

```{r}
shapiro.test(clean_data$PRICE)
```

```{r}
lambda = 0.0
```

```{r}
# Plot histogram for log-transformed response variable `PRICE`
hist(powerfun(clean_data$PRICE, lambda), col=adjustcolor('firebrick',alpha=0.6), 
     xlab='PRICE($)', ylab='Frequency',
     main='Histogram of Log-Transformed PRICE')
```

```{r}
shapiro.test(powerfun(clean_data$PRICE, lambda))
```

```{r}
trans_PRICE = powerfun(clean_data$PRICE, lambda)
```

```{r}
par(mfrow = c(2, 2))

plot(clean_data$BATHRM, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "BATHRM", ylab = "trans_PRICE",
     main = "BATHRM vs trans_PRICE")

plot(clean_data$HF_BATHRM, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "HF_BATHRM", ylab = "trans_PRICE",
     main = "HF_BATHRM vs trans_PRICE")

plot(clean_data$ROOMS, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "ROOMS", ylab = "trans_PRICE",  
     main = "ROOMS vs trans_PRICE")

plot(clean_data$BEDRM, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "BEDRM", ylab = "trans_PRICE", 
     main = "BEDRM vs trans_PRICE")
```

```{r}
par(mfrow = c(2, 2))

plot(clean_data$AYB, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "AYB", ylab = "trans_PRICE",
     main = "AYB vs trans_PRICE")

plot(clean_data$YR_RMDL, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "YR_RMDL", ylab = "trans_PRICE",
     main = "YR_RMDL vs trans_PRICE")

plot(clean_data$EYB, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "EYB", ylab = "trans_PRICE",  
     main = "EYB vs trans_PRICE")

plot(clean_data$SALEDATE, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "SALEDATE", ylab = "trans_PRICE", 
     main = "SALEDATE vs trans_PRICE")
```

```{r}
par(mfrow = c(2, 2))

plot(clean_data$KITCHENS, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "KITCHENS", ylab = "trans_PRICE",
     main = "KITCHENS vs trans_PRICE")

plot(clean_data$FIREPLACES, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "FIREPLACES", ylab = "trans_PRICE",
     main = "FIREPLACES vs trans_PRICE")

plot(clean_data$STORIES, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "STORIES", ylab = "trans_PRICE",  
     main = "STORIES vs trans_PRICE")

plot(clean_data$AC, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "AC", ylab = "trans_PRICE", 
     main = "AC vs trans_PRICE")
```

```{r}
par(mfrow = c(2, 2))

plot(clean_data$SALE_YEAR, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "SALE_YEAR", ylab = "trans_PRICE",
     main = "SALE_YEAR vs trans_PRICE")

plot(clean_data$SALE_AYB_DIFF, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "SALE_AYB_DIFF", ylab = "trans_PRICE",
     main = "SALE_AYB_DIFF vs trans_PRICE")

plot(clean_data$SALE_EYB_DIFF, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "SALE_EYB_DIFF", ylab = "trans_PRICE",  
     main = "SALE_EYB_DIFF vs trans_PRICE")

plot(clean_data$SALE_RMDL_DIFF, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "SALE_RMDL_DIFF", ylab = "trans_PRICE", 
     main = "SALE_RMDL_DIFF vs trans_PRICE")
```

```{r}
par(mfrow = c(1, 2))

plot(clean_data$GBA, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "GBA", ylab = "trans_PRICE",
     main = "GBA vs trans_PRICE")

plot(clean_data$LANDAREA, trans_PRICE,
     pch = 19, col = adjustcolor('firebrick', alpha = 0.1), 
     xlab = "LANDAREA", ylab = "trans_PRICE",
     main = "LANDAREA vs trans_PRICE")
```

# Model Building and Analysis

## Model Training
```{r}
library(glmnet)

X_train = model.matrix(powerfun(PRICE, lambda)~., data=data_train)[,-1]
X_test = model.matrix(powerfun(PRICE, lambda)~., data=data_valid)[,-1]
y_train = powerfun(data_train$PRICE, lambda)
y_test = powerfun(data_valid$PRICE, lambda)
```

```{r}
cv_lasso = cv.glmnet(X_train, y_train, alpha=1)
best_lasso_lambda = cv_lasso$lambda.min
```

```{r}
lasso_model = glmnet(X_train, y_train, alpha=1, lambda=best_lasso_lambda)
```

```{r}
# Check non-zero LASSO coefficients
predict(lasso_model, type="coefficients", s=best_lasso_lambda)
```

## Compute Loss Metrics

```{r}
# Predicted values are transformed by powerfun(PRICE, lambda)
valid_pred = predict(lasso_model, s=best_lasso_lambda, newx=X_test)

# Using inv_powerfun to convert back to original scale
inv_valid_pred = inv_powerfun(valid_pred, lambda)
```

### Compute MSE, RMSE and RMSLE

```{r}
# Compute metrics on validation dataset
mse = mean((data_valid$PRICE - inv_valid_pred)^2)
rmse = sqrt(mse)
rmsle = sqrt(mean((log(data_valid$PRICE) - log(inv_valid_pred))^2))
```

```{r}
cat("MSE:", mse, "\n")
cat("RMSE:", rmse, "\n")
cat("RMSLE:", rmsle, "\n")
```





